Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-17 02-antecedentes.qmd
reestructuracion antecedentes

Whether inequalities in education are considered just or not lead us almost naturally to the concepts of merit and meritocracy. Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.

@@

Our second argument is based on the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019]. Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps. In line with this evidence, we expect that:

$H_2$: Those with lower status characteristics show lower justification of educational inequalities

Thirdly, we address inequality perception. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.

Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:

@@

Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed. As a consequence, the justification of inequality could show a decrease over time:

@@

>>>>

@@

Besides proposing a general decrease over time in the justification of inequalities, a second hypothesis refers to specific aspects of the pandemic. As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, bonuses, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks. As a consequence, preferences for market justice would be reduced:

$H_2$: Recipient of state benefits show lower market justice preferences

In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger market justice preferences:

$H_3$: Those prioritizing health instead of the economic situation show higher market justice preferences

Whether inequalities in education are considered just or not lead us to take into account the concepts of merit and meritocracy. Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort. In line with this evidence, we propose that

@@

Combining the arguments regarding the effect of the pandemic and its relationship with meritocracy, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed. As a consequence, the justification of inequality could show a decrease over time:

@@

Along with meritocratic perceptions, in this study we incorporate inequality perception. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

$H_6$: A larger inequality perception is negatively associated with the justification of educational inequalities.

Next, we take into account the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019]. Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps. In line with this evidence, we expect that:

$H_7$: Those with lower status characteristics show lower justification of educational inequalities
